---
title: "Joining and cleaning data"
output: html_notebook
---

This notebook is based on the vignette about dplyr, and [this blog post about dplyr and tidyr](https://www.guru99.com/r-dplyr-tutorial.html).

To explore the basic data manipulation verbs of dplyr, we’ll use nycflights13::flights. This dataset contains all 336776 flights that departed from New York City in 2013. 


```{r}
vignette("two-table",package="dplyr")
library(dplyr)
library(nycflights13)
```


## Two Table verbs

It’s rare that a data analysis involves only a single table of data. In practice, you’ll normally have many tables that contribute to an analysis, and you need flexible tools to combine them. In dplyr, there are three families of verbs that work with two tables at a time:

* Mutating joins, which add new variables to one table from matching rows in another.

* Filtering joins, which filter observations from one table based on whether or not they match an observation in the other table.

* Set operations, which combine the observations in the data sets as if they were set elements.

All two-table verbs work similarly. The first two arguments are x and y, and provide the tables to combine. The output is always a new table with the same type as x.

### Mutating joins
Mutating joins allow you to combine variables from multiple tables. For example, take the nycflights13 data. In one table we have flight information with an abbreviation for carrier, and in another we have a mapping between abbreviations and full names. You can use a join to add the carrier names to the flight data:

```{r}
# Drop unimportant variables so it's easier to understand the join results.
flights2 <- flights %>% select(year:day, hour, origin, dest, tailnum, carrier)

# The airlines dataframe contains columns "carrier" and "name"
# Because flights2 also has a column named "carrier", it is  
# automatically used as a join column:

flights2 %>% 
  left_join(airlines)
# you might need to click on the right arrow in te result to show the names:
```

###Controlling how the tables are matched

As well as x and y, each mutating join takes an argument by that controls which variables are used to match observations in the two tables. There are a few ways to specify it, as I illustrate below with various tables from nycflights13:

* NULL, the default. dplyr will will use all variables that appear in both tables, a natural join. For example, the flights and weather tables match on their common variables: year, month, day, hour and origin.

```{r}
flights2 %>% left_join(weather)
```

* A character vector, by = "x". Like a natural join, but uses only some of the common variables. For example, flights and planes have year columns, but they mean different things so we only want to join by tailnum.

```{r}
flights2 %>% left_join(planes, by = "tailnum")
```

Note that the year columns in the output are disambiguated with a suffix.

* A named character vector: by = c("x" = "a"). This will match variable x in table x to variable a in table b. The variables from use will be used in the output. Note that is probably the safest way to create joins.

Each flight has an origin and destination airport, so we need to specify which one we want to join to:

```{r}
# let's first look at columns of interest within airports
airports %>% select(faa, name) %>% head()
```

```{r}
# now create the join for the dest column
flights2 %>% left_join(airports, c("dest" = "faa"))
```

```{r}
# or for the origin column:
flights2 %>% left_join(airports, c("origin" = "faa"))
```

###Types of join
There are four types of mutating join, which differ in their behaviour when a match is not found. We’ll illustrate each with a simple example:
```{r}
df1 <- tibble(x = c(1, 2), y = 2:1)
df2 <- tibble(x = c(1, 3), a = 10, b = "a")
df1
df2
```

* inner_join(x, y) only includes observations that match in both x and y.
```{r}
df1 %>% inner_join(df2)
```
* left_join(x, y) includes all observations in x, regardless of whether they match or not. This is the most commonly used join because it ensures that you don’t lose observations from your primary table.

```{r}
df1 %>% left_join(df2)
# all rows of df1 are shown, 
# columns of non matching rows in df2 are shown with NA values
```
* right_join(x, y) includes all observations in y. It’s equivalent to left_join(y, x), but the columns will be ordered differently.

```{r}
df1 %>% right_join(df2)
# all rows of df2 are shown, 
# columns of non matching rows in df1 are shown with NA values
```
* full_join() includes all observations from x and y.
```{r}
df1 %>% full_join(df2)
```


#Data Cleaning functions
Following are four important functions to tidy the data:

* gather(): Transform the data from wide to long
* spread(): Transform the data from long to wide
* separate(): Split one variable into two
* unit(): Unit two variables into one

We use the tidyr library. This library belongs to the collection of the library to manipulate, clean and visualize the data. 

```{r}
#install.packages("tidyr")
library(tidyr)
```

###gather()
The objectives of the gather() function is to transform the data from wide to long.

```{r}
library(tidyr)
# Create a messy dataset
messy <- data.frame(
  country = c("A", "B", "C"),
  q1_2017 = c(0.03, 0.05, 0.01),
  q2_2017 = c(0.05, 0.07, 0.02),
  q3_2017 = c(0.04, 0.05, 0.01),
  q4_2017 = c(0.03, 0.02, 0.04))
messy
```
As you can see, in this messy data frame there are multiple columns about the results of a quarter. The gather() function can be used to combine these columns into one column containing the names q1_2017 to q4_2017, and another column containing the values:

```{r}
# Reshape the data
tidier <-messy %>%
gather(quarter, growth, q1_2017:q4_2017)
tidier
```
###spread()
The spread() function does the opposite of gather. It accepts a dataframe, a key column and a value column, and will create a column for each key, containing the corresponding values.

We can reshape the tidier dataset back to messy with spread()

```{r}
# Reshape the data: quarter contains the keys, growth contains the values
messy_1 <- tidier %>%
  spread(quarter, growth) 
messy_1
```

###separate()
The separate() function splits a column into two according to a separator. This function is helpful in some situations where the variable is a date. Our analysis can require focussing on month and year and we want to separate the column into two new variables.

We can split the quarter from the year in the tidier dataset by applying the separate() function.

```{r}
separate_tidier <-tidier %>%
separate(quarter, c("Qrt", "year"), sep ="_")
head(separate_tidier)
```

###unite()
The unite() function concanates two columns into one.

In the above example, we separated quarter from year. What if we want to merge them. We use the following code:

```{r}
unit_tidier <- separate_tidier %>%
  unite(Quarter, Qrt, year, sep ="_")
head(unit_tidier)
```

